Lifespan analysis

### 2023-07-10 stats/lifespan.Rmd
clean up, add Bayesian changepoint (Rbeast)

```{r load_packages, message=FALSE, warning=FALSE, include=FALSE}
library(tidyverse) # data handling etc.
library(tidybayes) # for stat_halfeye, add_predictive_draws
library(tidyboot) # for bootstrapped CIs
library(brms) # modeling
library(cmdstanr) # modeling
check_cmdstan_toolchain(fix = TRUE, quiet = TRUE)

@@

```{r prepare_modeldata}
mdata <- lifespan_data %>% 
  group_by(subj_id, age, age_centered) %>% 
  summarize(mean_dist = mean(click_dist_balloons, na.rm = TRUE))
```

@@

```{r models, message=FALSE, warning=FALSE}
# linear age
mlinear <- brm(mean_dist ~ age_centered,
                    data = mdata,
                    family = lognormal(),
                    warmup = 2000, 
                    iter   = 6000, 
                    chains = 4, 
                    cores  = 4, 
                    threads = threading(8), 
                    control = list(adapt_delta = 0.95), 
                    backend = "cmdstanr") %>% add_criterion(c("loo","waic"))

@@

## quadratic age
mquadratic <- brm(mean_dist ~ 1 + age_centered + I(age_centered^2),
                    data = mdata,
                    family = lognormal(),
                    warmup = 2000, 
                    iter   = 6000, 
                    chains = 4, 
                    cores  = 4, 
                    threads = threading(8), 
                    control = list(adapt_delta = 0.95), 
                    backend = "cmdstanr") %>% add_criterion(c("loo","waic"))

@@

## cubic age
mcubic <- brm(mean_dist ~ 1 + age_centered + I(age_centered^2) + I(age_centered^3),
                    data = mdata,
                    family = lognormal(),
                    warmup = 2000, 
                    iter   = 6000, 
                    chains = 4, 
                    cores  = 4, 
                    threads = threading(8), 
                    control = list(adapt_delta = 0.95), 
                    backend = "cmdstanr") %>% add_criterion(c("loo","waic"))

@@

## gaussian process model
mgaussianprocess <- brm(mean_dist ~ gp(age_centered, k=50, c=5/4, scale=TRUE),
                    data = mdata,
                    family = lognormal(),
                    warmup = 2000, 
                    iter   = 6000, 
                    chains = 4, 
                    cores  = 4, 
                    threads = threading(8), 
                    control = list(adapt_delta = 0.95), 
                    backend = "cmdstanr") %>% add_criterion(c("loo","waic"))

mgaussianprocess

@@

```{r lifespan_modelcomparison}
# mlinear <- readRDS("../saves/lifespan_mlinear.rds")
# mquadratic <- readRDS("../saves/lifespan_mquadratic.rds")
# mcubic <- readRDS("../saves/lifespan_mcubic.rds")
# mgaussianprocess <- readRDS("../saves/lifespan_mgaussianprocess.rds")

@@

```{r draws_winning_model}
# mgaussianprocess <- readRDS("../saves/lifespan_mgaussianprocess.rds")

mdraws <- mdata %>%
  add_predicted_draws(mgaussianprocess) %>% 
  ungroup() %>% 
  select(-c(.row, .chain, .iteration)) %>%
  rename(draw = .draw, predicted_value = .prediction)

write.csv(mdraws, "../saves/lifespan_mdraws.csv", quote = F, row.names = F)
saveRDS(mdraws, "../saves/lifespan_mdraws.rds")
```

```{r prediction_plot_winning_model}
### this sections takes a while to run

# mdraws <- readRDS("../saves/lifespan_mdraws.rds")

mdraws %>%
  ggplot(aes(x = age)) +
  stat_lineribbon(aes(y = predicted_value), .width = c(.95), colour = "#006c66", fill = "lightgrey") +
  geom_jitter(aes(y = mean_dist), colour = "darkgrey", size = 1.5, alpha = 0.8) +
  scale_x_continuous(limits = c(0,80), breaks = seq(0, 80, 5)) +
  scale_y_continuous(limits = c(0,4)) +
  labs(x = "Age in years", y = "Mean click distance in target widths")

# saveRDS(prediction_plot, file = "../saves/lifespan_modelplot.rds")
ggsave("../figures/lifespan_modelplot.png", width = 8, height = 5, scale = 1, bg = "white")
```

```{r playground}
# mgaussianprocess <- readRDS("../saves/lifespan_mgaussianprocess.rds")

# model checks: 
# density plot overlay of the original response values (the thick black line) with 10 draws from the posterior distribution of the model
pp_check(mgaussianprocess)

# empirical cumulative distribution function of the observations and random draws from the model posterior:
pp_check(mgaussianprocess, type = "ecdf_overlay")

# plots distribution of estimates & chains converging
plot(mgaussianprocess)

# model plotting
spline_plot_data <- fitted(mgaussianprocess) %>% 
  data.frame() %>% 
  bind_cols(mdata)

spline_plot_data %>% 
  ggplot(aes(x = age, y = mean_dist, ymin = Q2.5, ymax = Q97.5)) + 
  geom_hline(yintercept = fixef(mgaussianprocess)[1, 1], linetype = 2) +
  geom_point(color = "steelblue") +
  geom_ribbon(alpha = 2/3) +
  labs(x = "Age",
       y = "Imprecision") +
  theme_classic()

# predictions & actual raw data underneath
conditional_effects(mgaussianprocess, ndraws = 200, spaghetti = TRUE) %>% 
  plot(points = T)

# get intercepts
posterior_summary(mgaussianprocess)
```

```{r smooths}
lifespan_smooths <- mdata %>%
  ggplot(aes(x = age, y = mean_dist)) +  
  geom_jitter(colour = "grey", size = 1.5, alpha = 0.8) + 
  geom_smooth(method = lm, se = F, aes(colour = "A")) +
  geom_smooth(method = lm, formula = y ~ poly(x, 2), se = F, aes(colour = "B")) +
  geom_smooth(method = lm, formula = y ~ poly(x, 3), se = F, aes(colour = "C")) +
  geom_smooth(method = loess, se = F, aes(colour = "D")) +
  geom_smooth(method = lm, formula = y ~ splines::bs(x, 5), se = F, aes(colour = "E")) +
  geom_smooth(method = gam, formula = y ~ s(x), se = F, aes(colour = "F")) +
  scale_fill_brewer(palette = "Greys") + 
  scale_x_continuous(limits = c(0,80), breaks = seq(0, 80, 5)) + 
  scale_y_continuous(limits = c(0,4)) + 
  scale_color_manual(name = "Smooth method", 
                     breaks = c("A", "B", "C", "D", "E", "F"),
                     values = c("#c6d325", "#00b1ea", "#29485d", "#9D3BBD", "#F71C8D", "orange"),
                     labels = c("linear", "quadratic", "cubic", "loess", "splines (k=5)", "splines (default)")) +
  labs(x = "Age in years", y = "Mean click distance in target widths")

lifespan_smooths

saveRDS(lifespan_smooths, file = "../saves/lifespan_smooths.rds")
ggsave("../figures/lifespan_smooths.png", width = 10, height = 6, scale = 0.7, bg = "white")
```

>>>>

```{r load_packages, message=FALSE, warning=FALSE, include=FALSE}
library(tidyverse) # data handling etc.
library(tidybayes) # for stat_halfeye, add_predictive_draws
library(tidyboot) # for bootstrapped CIs
library(brms) # modeling
library(cmdstanr) # modeling
library(Rbeast) # Bayesian change point analysis
check_cmdstan_toolchain(fix = TRUE, quiet = TRUE)

@@

```{r prepare_modeldata}
mdata <- lifespan_data %>% 
  group_by(subj_id, age, age_centered) %>% 
  summarize(mean_imprecision = mean(click_dist_balloons, na.rm = TRUE))
```

@@

# MODEL COMPARISON FOR DEVELOPMENTAL TRAJECTORY

```{r models, message=FALSE, warning=FALSE}
# linear age
mlinear <- brm(mean_imprecision ~ age_centered,
                    data = mdata,
                    family = lognormal(),
                    warmup = 2000, 
                    iter   = 6000, 
                    chains = 4, 
                    cores  = 4, 
                    threads = threading(8), 
                    control = list(adapt_delta = 0.95), 
                    backend = "cmdstanr") %>% add_criterion(c("loo","waic"))

@@

## quadratic age
mquadratic <- brm(mean_imprecision ~ 1 + age_centered + I(age_centered^2),
                    data = mdata,
                    family = lognormal(),
                    warmup = 2000, 
                    iter   = 6000, 
                    chains = 4, 
                    cores  = 4, 
                    threads = threading(8), 
                    control = list(adapt_delta = 0.95), 
                    backend = "cmdstanr") %>% add_criterion(c("loo","waic"))

@@

## cubic age
mcubic <- brm(mean_imprecision ~ 1 + age_centered + I(age_centered^2) + I(age_centered^3),
                    data = mdata,
                    family = lognormal(),
                    warmup = 2000, 
                    iter   = 6000, 
                    chains = 4, 
                    cores  = 4, 
                    threads = threading(8), 
                    control = list(adapt_delta = 0.95), 
                    backend = "cmdstanr") %>% add_criterion(c("loo","waic"))

@@

## gaussian process model
mgaussianprocess <- brm(mean_imprecision ~ gp(age_centered, k=50, c=5/4, scale=TRUE),
                    data = mdata,
                    family = lognormal(),
                    warmup = 5000, 
                    iter   = 20000, 
                    chains = 4, 
                    cores  = 4, 
                    threads = threading(8), 
                    control = list(adapt_delta = 0.95), 
                    backend = "cmdstanr")

mgaussianprocess <- mgaussianprocess %>% add_criterion(c("loo","waic"))

@@

```{r modelcomparison}
# mlinear <- readRDS("../saves/lifespan_mlinear.rds")
# mquadratic <- readRDS("../saves/lifespan_mquadratic.rds")
# mcubic <- readRDS("../saves/lifespan_mcubic.rds")
# mgaussianprocess <- readRDS("../saves/lifespan_mgaussianprocess.rds")

@@

```{r draws_winningmodel}
# mgaussianprocess <- readRDS("../saves/lifespan_mgaussianprocess.rds")

# mdraws <- mdata %>%
#   add_predicted_draws(mgaussianprocess) %>% 
#   ungroup() %>% 
#   select(-c(.row, .chain, .iteration)) %>%
#   rename(draw = .draw, predicted_value = .prediction)
# 
# write.csv(mdraws, "../saves/lifespan_mdraws.csv", quote = F, row.names = F)
# saveRDS(mdraws, "../saves/lifespan_mdraws.rds")

nd <- mdata %>% 
  select(-c(mean_imprecision))

pred <- fitted(mgaussianprocess, newdata = nd, re_formula = NA) %>% as_tibble() %>% bind_cols(nd) 
```

# CHANGE POINT ANALYSIS

```{r changepoints}
x <- mdata %>% arrange(age) %>% pull(age)
y <- mdata %>% arrange(age) %>% pull(mean_imprecision)

## irregular time intervals
ibeast <- beast.irreg(y = y, time = x, season = "none", 
            torder.minmax = c(0, 0), # for constant
            tcp.minmax = c(0, 5), # we assume 3, allow some room for exploration
            tseg.min = 10, 
            mcmc.chains = 4, mcmc.burnin = 1000, mcmc.samples = 10000, 
            ci = TRUE)

print(ibeast)
plot(ibeast)

ibeast$trend$cp
```

# PLOT

```{r plot}
ggplot() +

  geom_jitter(
    data = mdata, 
    aes(x = age, y = mean_imprecision), 
    col = "darkgrey", alpha = 0.75
  ) +

  geom_smooth(
    data = pred, 
    aes(x = age, y = Estimate, ymin = Q2.5, ymax = Q97.5), 
    stat = "identity", size = 1, alpha = 0.15, fill = "#006C66", col = "#006C66",
  ) +

  # TODO: add lines where changepoints
  scale_x_continuous(limits = c(0, 80), breaks = seq(0, 80, 5)) + 
  scale_y_continuous(limits = c(0, 4), breaks = seq(0, 4, 0.5)) +

  labs(x = "Age in years", y = "Average imprecision in target width")
```
